Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based cohort study

### 2024-06-26 manuscript2.Rmd
26/6

---
title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
output: "html_document"
author: "Elin Sun Cao"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options: 
  markdown: 
    wrap: 72
---

## Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study
### Background
Trauma is a leading cause of mortality and morbidity worldwide. A cornerstone in trauma quality improvement programs is multidisciplinary mortality and morbidity reviews. The purpose of these reviews is to identify opportunities for improvement and corresponding corrective actions. Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this cohort. The aim of this study was to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.

@@

# Introduction
Trauma — the clinical entity of external injury and the body’s associated response — is a leading cause of mortality and morbidity worldwide [@60; @6]. Patients with major trauma often require intensive care unit (ICU) admission, where managing these patients can be challenging and can result in complications or death [@64, @69]. Although these complications are often expected, the complexity of ICU care necessitates providers to be highly adaptable, which makes the setting prone to mistakes and errors. Peer review is a cornerstone of trauma quality improvement programs aimed at identifying preventable causes of mortality and morbidity in trauma patients [@70], referred to as opportunities for improvement [@42]. The identification of these help derive corrective action plans to improve the quality of trauma care. Previous research on opportunities for improvement in trauma care have been compiled in a systematic review, which classified and identified areas for improvement such as inadequate monitoring, team communication, assessment of injury and procedure-related issues [@75].

Assessing preventable deaths among ICU-admitted patients poses a significant challenge due to their medical complexity, making it difficult to establish the standard of care for each individual. Moreover, despite receiving optimal treatment, these patients remain at elevated risk for complications and increased mortality rates [@9]. While previous studies have explored various facets of trauma care, including the ICU-setting [@50], there is a notable gap in research regarding error reporting practices in trauma care - especially for trauma patients admitted to the ICU [@64; @12]. Understanding and addressing this gap is essential for improving patient outcomes and optimizing trauma care protocols. Therefore, this study aims to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.

@@

## Study design

We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, which is included in the Swedish Trauma registry, and the hospital's local trauma care quality database.

@@

The Karolinska University Hospital in Solna is the trauma center designated to receive all severely injured patients in the wider metropolitan area of Stockholm [@17]. The hospital has direct access to radiology, surgery, intensive care, and consultants from relevant specialities [@18]. The care of adult trauma patients is screened for opportunities for improvement using a combination of audit filters and individual review by specialized nurses. Those identified as having a higher likelihood of opportunities for improvement are discussed in a multidisciplinary peer-review held every 6-8 weeks. The presence or absence of opportunities for improvement is decided among the participants of the review and subsequently documented in the trauma care quality database. Prior to 2017 a semi-systematic selection off clinicians involved in the opportunities for improvement process was done by filtering for patients who were thought to have a high risk for opportunities for improvement. However, starting in 2017, the care of all patients is screened using audit filters.

Table 1 illustrates opportunities for improvement divided into categories based on the definitions of opportunity for improvement types during mortality and morbidity reviews at the Karolinska University Hospital in Solna.

@@

When conducting logistic regression analysis, the number of events per variable equal to 10 is a rule of thumb. If the events per variable falls below this threshold, the results for regression analyses should be interpreted with caution [@45].

@@

We categorized and converted the independent variables according to the aforementioned guideline of events per variable to achieve an acceptable multivariate analysis. Specifically, Revised Trauma Score, Injury Severity Score, and time to first computed tomography were treated as continuous variables, while categorical variables were divided into as few levels as possible.

@@

The variables selected for analysis were patient and process factors chosen from the trauma registry, based on the locally used audit filters, previous literature and expert opinion. Categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, ASA preinjury, and 30-day mortality. Sex was categorized as male sex or female sex. Mechanical ventilation was stratified into “Not intubated”, “Mechanical ventilation 1-7 days”, “Mechanical ventilation > 7 days” and “Unknown”. On-call hours was defined as arrival to the hospital before 8.00 a.m., after 5 p.m., or during weekends – Saturday or Sunday. ICU length of stay was categorized as “≤ 7 days” or “> 7 days”. American Society of Anesthesiologists (ASA) preinjury was divided into “ASA 1-2” and “ASA 3-6”. 30-day mortality was determined based on whether the patient was deceased or alive 30 days after injury.

The continuous variables in the analysis were age in years, Revised Trauma Score, Injury Severity Score, and time to first computed tomography in minutes. Revised Trauma Score is a measure of severity based on prehospital or emergency department values. In instances where hospital values for systolic blood pressure, respiratory rate, or Glasgow Come Scale scores were missing from the emergency department records, prehospital values were used instead. If the patient was intubated, it was assumed that their Glasgow Coma Scale was three and respiratory rate was zero. Subsequently, for the Revised Trauma Score variable, each of the variables Glasgow Coma Scale, systolic blood pressure, and respiratory rate were converted to Revised Trauma Score codes ranging from 0 to 4 based on their respective values, and the final score was calculated using published coefficients [@71].

### Study outcome

The study’s outcome was the presence of any opportunity for improvement. An opportunity for improvement presents preventable errors in care with direct adverse effects on patient outcomes or recurrent deviations from safe clinical practice. Examples include technical error, clinical judgement error or any preventable death.

@@

We conducted complete case analysis after handling missing values for systolic blood pressure, respiratory rate and Glasgow Coma Scale as described above. Pearson’s Chi-squared test and the Wilcoxon rank sum test were used to calculate the significance for categorical and continuous variables respectively. Continuous variables are presented using mean values and standard deviations, and categorical variables are presented with the number of patients in each subgroup and their respective percentages.

Univariate logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and opportunities for improvement. Odds ratios (OR) with associated 95% confidence intervals were calculated. A p-value of less than 0.05 considered indicative of a significant difference. For continuous variables, ORs were derived by evaluating each unitary step increment. A subgroup analysis of patients who were alive and dead 30 days after the injury was conducted to reduce impact from death as a cause of the association between opportunities for improvement and patient and process factors. All statistical analyses were conducted using R version 4.4 [@19].

@@

The patient demographics of this study is presented in Figure 1. A total of 14,022 patients were included in the trauma registry and trauma care quality database between 2014 and 2023, out of which 12,278 patients were excluded. Out of all patients, 11,343 were excluded because they were not treated in the ICU, three patients were excluded because they were under 15 years old, six patients were excluded because they were dead on arrival, and 928 patients were excluded because they had not been screened regarding the presence of opportunities for improvement This left us with a total of 1,742 patients eligible for the study. Out of these, 293 patients were excluded because there was missing data in one or several of the patient and process factors, leaving a total of 1,449 patients included in the study. The variable with the highest number of missing values was Revised Trauma Score, with 207 instances of missing data.

@@

Table 2 present sample characteristics. Most patients were male 1,096 (76%) vs 353 (24%). There were 846 (58%) patients who were not intubated, 241 (17%) patients who received mechanical ventilation for 1-7 days, and 113 (7,8%) patients who received mechanical ventilation for more than 7 days. The number of patients arriving outside working hours was 1,042 (72%). A total of 1,064 (73%) patients were equal to or less than 7 days in the ICU vs 385 (27%) patients whose ICU-stay exceeded 7 days. A total of 1,080 (75%) patients had an ASA score of 1-2 before arrival to the hospital, and 369 (25%) patients had an ASA score of 3-6 before arrival to the hospital. Amongst our patients, 30-day mortality was at 286 (20%).

When comparing patients with opportunities for improvement to those without, differences emerged across various factors. Specifically, individuals with opportunities for improvement were less likely to require intubation, had higher Revised Trauma Score, experienced shorter time to initial computed tomography scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.

@@

Table 3 shows the unadjusted and adjusted associations between patient and process factors and opportunities for improvement. In the unadjusted analysis, the factors that were significantly associated with the presence of opportunities for improvement were: Revised Trauma Score `r inline_text(table3a, variable = RTS)`, > 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3a, variable = Survival, level = "Alive")`.

In the adjusted analysis the following factors demonstrated significant association with opportunities for improvement: Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`.

Sex, age, mechanical ventilation, Injury Severity Score, time to first computed tomography, on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in the unadjusted nor the adjusted analyses.

@@

Table 4 shows the adjusted associations of opportunities for improvement and the selected patient and process factors among the patients who were alive and dead respectively 30 days after injury. Revised Trauma Score was significantly associated with increased odds of OFI for both patients who were alive `r inline_text(table3aalive, variable = RTS)` and deceased `r inline_text(table3bdead, variable = RTS)` 30 days after injury.

@@

The key results from this study are that higher Revised Trauma Score, > 7 days length of stay in the ICU, and survival 30 days after injury were significantly associated with higher odds of opportunities for improvement. Age, sex, mechanical ventilation, Injury Severity Score, time to first computed tomography, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in neither the adjusted nor the unadjusted analyses. In the analysis of patients who were alive and deceased 30 days after injury increased Revised Trauma Score was significantly associated with higher odds of opportunities for improvement in both patient groups. Among the patients who deceased 30 days after injury, more than 7 days in the ICU was significantly associated with higher odds of opportunities for improvement.

Our study found a significant association between higher Revised Trauma Score and higher odds of opportunities for improvement, indicating that among very severely injured patients that are admitted to the ICU, patients in a less severe condition were at higher odds of opportunities for improvement. Patients with lower Revised Trauma Score are often in critical condition upon arrival — upon which emergency interventions are performed immediately. In contrast, patients with higher Revised Trauma Score, indicating relative stability upon arrival, may experience delays in interventions overall and might receive less rigorous monitoring. This aligns with findings from previous research conducted on trauma patients from the Karolinska University Hospital Solna, where it was noted that patients with more stable vital parameters were more likely to have increased odds of opportunities for improvement [@12].

>>>>

---
title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based cohort study"
output: "html_document"
author: "Elin Sun Cao"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options: 
  markdown: 
    wrap: 72
---

## Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based cohort study
## Abstract
### Background
Trauma is a leading cause of mortality and morbidity worldwide. A cornerstone in trauma quality improvement programs is multidisciplinary mortality and morbidity reviews. The purpose of these reviews is to identify opportunities for improvement and corresponding corrective actions. Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this cohort. The aim of this study was to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.

@@

# Introduction
Trauma — the clinical entity of external injury and the body’s associated response — is a leading cause of mortality and morbidity worldwide [@60; @6]. Patients with major trauma often require intensive care unit (ICU) admission, where managing these patients can be challenging and result in complications or death [@64; @69]. Although these complications are often expected, the complexity of ICU care makes the setting prone to mistakes and errors.

Peer review is a cornerstone of trauma quality improvement programs aimed at identifying preventable causes of mortality and morbidity in trauma patients [@70], referred to as opportunities for improvement [@42]. Previous research on opportunities for improvement in trauma care has focused on the initial resuscitation phase and show that
*detta ska redigeras*
been compiled in a systematic review, which classified and identified areas for improvement such as inadequate monitoring, team communication, assessment of injury and procedure-related issues [@75].

In contrast, little is known about opportunities for improvement in the care of trauma patients in the ICU [@64; @12]. Assessing opportunities for improvement in these patients is challenging because it is difficult to establish the standard of care for each individual and despite optimal treatment, these patients are at risk for complications and death [@9].  This study aims to characterize opportunities for improvement in adult trauma patients admitted to the ICU and assess how patient and process factors are associated with opportunities for improvement in these patients.

@@

## Study design and data sources

We conducted a retrospective single-center cohort study based on data from the Karolinska University Hospital trauma registry, and the hospital's local trauma care quality database. The trauma registry is part of the Swedish national trauma registry SweTrau [*källa*].

@@

The Karolinska University Hospital in Solna is the trauma center designated to receive all severely injured patients in the wider metropolitan area of Stockholm [@17]. The hospital has direct access to radiology, surgery, intensive care, and consultants from relevant specialities [@18]. The care of adult trauma patients is screened for opportunities for improvement using a combination of audit filters and individual review by specialized nurses. Those identified as having a higher likelihood of opportunities for improvement are discussed in a multidisciplinary peer-review meeting held every 6-8 weeks. The presence or absence of opportunities for improvement is decided among the participants of the review and subsequently documented in the trauma care quality database.

@@

We included all eligible patients from the trauma registry.

@@

### Outcome

The outcome was the presence of any opportunity for improvement. An opportunity for improvement presents preventable events in the care of patients with adverse effects on their outcomes, or recurrent deviations from safe clinical practice.

@@

The variables selected for analysis were patient and process factors chosen from the trauma registry, based on the locally used audit filters, previous literature and expert opinion. The categorical factors included sex, mechanical ventilation, on-call hours, ICU length of stay, American Society of Anesthesiologists preinjury score, and 30-day mortality. Sex was categorized as male sex or female sex. Mechanical ventilation was stratified into “Not intubated”, “Mechanical ventilation 1-7 days”, “Mechanical ventilation > 7 days” and “Unknown”. On-call hours was defined as arrival to the hospital before 8.00 a.m., after 5 p.m., or during weekends – Saturday or Sunday. ICU length of stay was categorized as “≤ 7 days” or “> 7 days”. American Society of Anesthesiologists (ASA) preinjury score was divided into “ASA 1-2” and “ASA 3-6”. We categorised variables to limit the number of independent variables in the multivariable analysis to adhere to the rule of thumb of ten patients with the outcome per variable [@45].

The continuous variables in the analysis were age in years, Revised Trauma Score, Injury Severity Score, and time to first computed tomography in minutes. The Revised Trauma Score is a measure of severity based on prehospital or emergency department values. In instances where hospital values for systolic blood pressure, respiratory rate, or Glasgow Come Scale scores were missing from the emergency department records, prehospital values were used instead. If the patient was intubated, it was assumed that their Glasgow Coma Scale was three and respiratory rate was zero. Subsequently, for the Revised Trauma Score variable, each of the variables Glasgow Coma Scale, systolic blood pressure, and respiratory rate were converted to Revised Trauma Score codes ranging from 0 to 4 based on their respective values, and the final score was calculated using published coefficients [@71].

@@

We conducted a complete case analysis after handling missing values for systolic blood pressure, respiratory rate and Glasgow Coma Scale as described above. Continuous variables are presented using mean values and standard deviations, and categorical variables are presented with the number of patients in each subgroup and their respective percentages.

We used the Pearson’s Chi-squared test and the Wilcoxon rank sum test to compare categorical and continuous variables respectively between patients with and without opportunities for improvement.

We used logistic regression to determine unadjusted and adjusted associations between patient and process factors and opportunities for improvement. Odds ratios (OR) with associated 95% confidence intervals were calculated. A p-value of less than 0.05 considered indicative of a significant difference. As a sensitivity analysis, we considered 30-day mortality as a potential effect modifier and therefore conducted a stratified analysis in which the associations between patient and process factors and opportunities for improvement were estimated separately in patients who survived and patients who died within 30 days. All statistical analyses were conducted using R version 4.4 [@19].

@@

A total of 14,022 patients were included in the trauma registry and trauma care quality database between 2014 and 2023, out of which 12,278 patients were excluded, as shown in Figure 1. Out of all patients, 11,343 were excluded because they were not treated in the ICU, three patients were excluded because they were under 15 years old, six patients were excluded because they were dead on arrival, and 928 patients were excluded because they had not been screened regarding the presence of opportunities for improvement This left us with a total of 1,742 patients eligible for the study. Out of these, 293 patients were excluded because there was missing data in one or several of the patient and process factors, leaving a total of 1,449 patients included in the study. The variable with the highest number of missing values was Revised Trauma Score, with 207 instances of missing data.

@@

Table 1 present sample characteristics. Most patients were male 1,096 (76%) vs 353 (24%). There were 846 (58%) patients who were not intubated, 241 (17%) patients who received mechanical ventilation for 1-7 days, and 113 (7,8%) patients who received mechanical ventilation for more than 7 days. The number of patients arriving outside working hours was 1,042 (72%). A total of 1,064 (73%) patients were equal to or less than 7 days in the ICU vs 385 (27%) patients whose ICU-stay exceeded 7 days. A total of 1,080 (75%) patients had an ASA score of 1-2 before arrival to the hospital, and 369 (25%) patients had an ASA score of 3-6 before arrival to the hospital. Amongst our patients, 30-day mortality was at 286 (20%). When comparing patients with opportunities for improvement to those without, patients with opportunities for improvement were less likely to require intubation, had higher Revised Trauma Score, experienced shorter time to initial computed tomography scans, spent more days in the ICU, and demonstrated a greater likelihood of surviving 30 days post-injury.

Table 2 illustrates opportunities for improvement divided into categories based on the definitions of opportunity for improvement types during mortality and morbidity reviews at the Karolinska University Hospital in Solna.

@@

Table 3 shows the unadjusted and adjusted associations between patient and process factors and opportunities for improvement. In the unadjusted analysis, Revised Trauma Score `r inline_text(table3a, variable = RTS)`, > 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3a, variable = Survival, level = "Alive")`, were all significantly associated with opportunities for improvement. These variables remained significantly associated with opportunities for improvement in the adjusted analysis, with similar effect magnitudes: Revised Trauma Score `r inline_text(table3b, variable = RTS)`, > 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "> 7 days")` and survival 30 days after injury `r inline_text(table3b, variable = Survival, level = "Alive")`. Sex, age, mechanical ventilation, Injury Severity Score, time to first computed tomography, on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in the unadjusted nor the adjusted analyses.

@@

Table 4 shows the results of the stratified analysis with adjusted associations between patient and process factors and opportunities for improvement among the patients who were alive and dead 30 days after injury respectively. The Revised Trauma Score was significantly associated with increased odds of OFI for both patients who were alive `r inline_text(table3aalive, variable = RTS)` and deceased `r inline_text(table3bdead, variable = RTS)` 30 days after injury.

@@

The key results from this study are that higher Revised Trauma Score, > 7 days length of stay in the ICU, and survival 30 days after injury were significantly associated with higher odds of opportunities for improvement. Age, sex, mechanical ventilation, Injury Severity Score, time to first computed tomography, arrival during on-call hours, and ASA preinjury were not significantly associated with opportunities for improvement in neither the unadjusted nor the adjusted analyses. In the stratified analysis of patients who were alive and dead 30 days after injury, only the Revised Trauma Score was significantly associated with higher odds of opportunities for improvement in both patient groups. Among the patients who deceased 30 days after injury, more than 7 days in the ICU was significantly associated with higher odds of opportunities for improvement.

Our study found a significant association between higher Revised Trauma Score and higher odds of opportunities for improvement, indicating that among very severely injured patients that are admitted to the ICU, patients in a less severe condition were at higher odds of opportunities for improvement. Patients with lower Revised Trauma Score are in a critical condition upon arrival — upon which emergency interventions are performed immediately. In contrast, patients with higher Revised Trauma Score, indicating relative stability upon arrival, may experience delays in interventions overall and might receive less rigorous monitoring. This aligns with findings from previous research conducted on trauma patients from the Karolinska University Hospital Solna, where it was noted that patients with more stable vital parameters were more likely to have opportunities for improvement [@12].
